---
title: "Police Killings and Violence on State Level"
author: "Babak Amini"
date: 05/24/2022
output: 
  html_document
---
```{r setup, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE}
knitr::opts_chunk$set(echo = FALSE)
```

**Links to spreadsheets**

[prisoner and crime in united state dataset](https://www.kaggle.com/christophercorrea/prisoners-and-crime-in-united-states)  
  
[police killings dataset](https://www.kaggle.com/mysarahmadbhat/police-killings)

**1. Introduction to Crime Datasets**  
  
This data is provided by the Bureau of Justice Statistics along with the FBI Uniform Crime Reporting Statistics. The belief is that locking away criminals with minimum sentencing has decreased violent crime since the 1980s. Therefore, the question is, are these high incarceration rates good for stopping crime, or are they counter productive? In the United States, one quarter of global prisoners are housed. This is obviously a high percentage given the United States is 4% of the global population.

In these datasets, we are looking at how violent crime relates to incarceration and police shootings. This is sparking interest as the death of George Floyd has come about. There are many different factors to study, but instead of making this political, the goal of these datasets is to see correlations between locking people in prison and if states with greater incarceration are incarcerating for violence or not. Additionally, are the states with more violence the states where there are a disproportionate number of police shootings? These questions have been on the public's mind for several years now, as the news media continues to spark outrage of police brutality.

**2. Questions to Look Into**  
  
1) When states have larger prison populations as a proportion of their state population, are the crimes disproportionately violent compared to states with less of their population incarcerated? This clarifies if states with more prisoners have more prisoners due to mass incarceration, or due to more violence.   
  
2) Do states with more violent crime per capita have more deaths from police shootings to stop crime?  
    
3) Does class status have a correlation with what weapons people carry when they are shot by police?  
  
4) What are the proportion of weapons that make up the most police shootings?  
  
5) How many people are killed by police from each race, and with what weapons were they killed? Are there any differences betweeen the race, and are those difference substantial?

```{r setUp, warning=FALSE, message=FALSE, echo = FALSE}
library('tidyverse')
library('ggplot2')
library('dbplyr')

stateData <- read.csv('./csv/crime_and_incarceration_by_state.csv')

prisonCustody <- read.csv('./csv/prison_custody_by_state.csv')

ucrByState <- read.csv('./csv/ucr_by_state.csv')

policeKillingsOrg <- read.csv('./csv/police_killings.csv')
policeKillings <- read.csv('./csv/police_killings.csv')

policeKillings$state[policeKillings$state == "AL"] <- "ALABAMA"
policeKillings$state[policeKillings$state == "AK"] <- "ALASKA"
policeKillings$state[policeKillings$state == "MA"] <- "MASSACHUSETTS"
policeKillings$state[policeKillings$state == "RI"] <- "RHODE ISLAND"
policeKillings$state[policeKillings$state == "LA"] <- "LOUSIANA"
policeKillings$state[policeKillings$state == "WI"] <- "WISCOUNSIN"
policeKillings$state[policeKillings$state == "CA"] <- "CALIFORNIA"
policeKillings$state[policeKillings$state == "AZ"] <- "ARIZONA"
policeKillings$state[policeKillings$state == "TX"] <- "TEXAS"
policeKillings$state[policeKillings$state == "MI"] <- "MICHIGAN"
policeKillings$state[policeKillings$state == "IN"] <- "INDIANA"
policeKillings$state[policeKillings$state == "TN"] <- "TENNESSEE"
policeKillings$state[policeKillings$state == "GA"] <- "GEORGIA"
policeKillings$state[policeKillings$state == "VA"] <- "VIRGINIA"
policeKillings$state[policeKillings$state == "NE"] <- "NEBRASKA"
policeKillings$state[policeKillings$state == "KS"] <- "KANSAS"
policeKillings$state[policeKillings$state == "OK"] <- "OKLAHOMA"
policeKillings$state[policeKillings$state == "OR"] <- "OREGON"
policeKillings$state[policeKillings$state == "NH"] <- "NEW HAMPSHIRE"
policeKillings$state[policeKillings$state == "FL"] <- "FLORIDA"
policeKillings$state[policeKillings$state == "IA"] <- "IOWA"
policeKillings$state[policeKillings$state == "NC"] <- "NORTH CAROLINA"
policeKillings$state[policeKillings$state == "NM"] <- "NEW MEXICO"
policeKillings$state[policeKillings$state == "ID"] <- "IDAHO"
policeKillings$state[policeKillings$state == "CO"] <- "COLORADO"
policeKillings$state[policeKillings$state == "HI"] <- "HAWAII"
policeKillings$state[policeKillings$state == "KY"] <- "KENTUCKY"
policeKillings$state[policeKillings$state == "SC"] <- "SOUTH CAROLINA"
policeKillings$state[policeKillings$state == "WV"] <- "WEST VIRGINIA"
policeKillings$state[policeKillings$state == "UT"] <- "UTAH"
policeKillings$state[policeKillings$state == "MO"] <- "MISSOURI"
policeKillings$state[policeKillings$state == "MD"] <- "MARYLAND"
policeKillings$state[policeKillings$state == "NY"] <- "NEW YORK"
policeKillings$state[policeKillings$state == "NJ"] <- "NEW JERSEY"
policeKillings$state[policeKillings$state == "IL"] <- "ILLINOIS"
policeKillings$state[policeKillings$state == "CT"] <- "CONNECTICUT"
policeKillings$state[policeKillings$state == "MS"] <- "MISSISSIPPI"
policeKillings$state[policeKillings$state == "NV"] <- "NEVADA"
policeKillings$state[policeKillings$state == "MT"] <- "MONTANA"
policeKillings$state[policeKillings$state == "MN"] <- "MINNESOTA"
policeKillings$state[policeKillings$state == "ME"] <- "MAINE"
policeKillings$state[policeKillings$state == "DE"] <- "DELAWARE"
policeKillings$state[policeKillings$state == "WY"] <- "WYOMING"
policeKillings$state[policeKillings$state == "OH"] <- "OHIO"
policeKillings$state[policeKillings$state == "WA"] <- "WASHINGTON"
policeKillings$state[policeKillings$state == "PA"] <- "PENNSYLVANIA"
policeKillings$state[policeKillings$state == "AR"] <- "ARKANSAS"

crimeAndPoliceInvolvement <- inner_join(stateData, policeKillings, by=c("jurisdiction" = "state", "year" = "year"))

view(stateData)
view(prisonCustody)
view(ucrByState)
view(policeKillings)
view(crimeAndPoliceInvolvement)
```

**3. Analysis**

```{r first, warning=FALSE, message=FALSE, echo = FALSE}
data <- stateData %>%
  mutate(percentagePopulation = (prisoner_count/state_population)*100,
         violentPercentage = violent_crime_total/(violent_crime_total+property_crime_total)*100) 

ggplot(data, aes(percentagePopulation, violentPercentage)) +
  geom_point() +
  geom_smooth(method=loess) +
  labs(x="Percentage of the State Population in Prison", y="Percentage of Crimes That are Violent", title="Violent Crime as a Percentage of All Crime as the Prisoner Count Increases") +
  theme(plot.title=element_text(hjust=.5, face="bold")) +
  scale_x_continuous(breaks=seq(-.05,1,by=.05)) +
  scale_y_continuous(breaks=seq(0,25,by=5))

mod <- lm(violentPercentage ~percentagePopulation, data = data)
```
  
In the above graph, we see how the percentage of the incarcerated for being violent increases, as the percentage of the state population in prison increases, at an increasing rate. This means that the states are focusing on violent crimes as they increase prison population. This suggests that the states with more prisoners are not rounding up people involved in no-victim crimes. If that was the case, we would see the violent crimes as a percentage of all incarcerated decrease as the prison population increases per state. Therefore, the more people are in prison, the more of them are incarcerated for violence, which is a healthy trend. However, this evidence does not suggest that the incarceration rate is justified. For instance, it is possible that a significant proportion of the violent crimes were self defense, and the state laws deemed them guilty for an arbitrary reason, such as a case where self-defense is morally justified but illegal because the prisoner did not attempt to retreat before using force. Another reason that this graph does not indicate if the incarceration rate is justified is because there is no data as to whether or not the increase in incarceration causes people to become violent. For instance, if an individual is incarcerated for marijuana possession, and after being incarcerated, they resort to robbery because they cannot get a job.  


```{r second, warning=FALSE, message=FALSE, echo = FALSE}
data <- crimeAndPoliceInvolvement %>%
  group_by(jurisdiction) %>%
  mutate(violentCrimeProp = 100*(violent_crime_total/state_population)) %>%
  select(jurisdiction, year, prisoner_count, state_population, violentCrimeProp, raceethnicity) %>%
  summarize(total = n(), percentPopShot = (total/state_population)*100, violentCrimeProp, raceethnicity) %>%
  arrange(desc(percentPopShot)) %>%
  distinct()

ggplot(data, aes(x=violentCrimeProp, y=percentPopShot)) +
  geom_point() + geom_smooth(method=lm, se=FALSE) +
  labs(x="Percentage of Violent Crime to State Population", y="Percent of Population Shot and Killed By Police", title="Correlation Between Police Shootings and State Violence (2015)") +
  theme(plot.title=element_text(hjust=.5, face="bold")) +
  scale_x_continuous(breaks=seq(0,1,by=.1)) +
  scale_y_continuous(breaks=seq(0,by=.0001))
```
  
  
The above figure shows a slight correlation between the amount of violent crime in a state adjusted for population, versus the number of police killings adjusted for population. As the figure shows there are several outliars in the data, with Oklahoma being the state with the most police killings per capita in 2015. Connecticut has the least percentage of their residents killed by police, with a grand total of 1. This result shows that police shootings have less of a relation to state violence than one could expect. It is possible that it has to do with the communities where these shootings are taking place, shown in the fourth graph, where most of the shootings took place in relatively low income areas. Therefore, the number of police shootings based on state violence, despite showing a correlation, is surprising.   


```{r third, warning=FALSE, message=FALSE, echo = FALSE}
policeKillingsOrg %>%
  group_by(state) %>%
  summarize(total=n(), state, Armed = armed) %>%
  ggplot() +
  geom_bar(aes(x=reorder(state, (total)), fill=Armed)) +
  scale_x_discrete(guide = guide_axis(n.dodge=2)) +
  coord_flip() +
  labs(x="State", y="Number of Police Shootins", title="Police Shootings Per State in 2015") +
  theme(plot.title=element_text(hjust=.5, face="bold")) +
  theme_dark() 
```
  
The above figure shows that police shootings are more or less related to population size, with California, Texas, and Florida in the top three for police shootings and also population, in the United States. The data shows that most of those shot and killed by police were armed with a firearm or knife, although the data does not specify if the weapons were legally carried or used during the shooting. About half who were shot, based on the above graph, were armed with a firearm, being the weapon of choice. Vehicles and Unknown arms make up a small percentage of each state's shootings. These bars suggest that most of those shot had a weapon, meaning that police were not shooting people just to shoot them. This suggests that the police feared for their lives or others when the shooting took place. The reason for this is that most people do not carry a weapon, so the disproportionate number of shooting victims with a weapon suggests the police did not randomly target these individuals. That is not to say that these shooting are justified or not justified, but to say that a potential for reasonable fear is plausible. Justification ought to be decided on a case by case basis.  

```{r fourth, warning=FALSE, message=FALSE, echo = FALSE}
policeKillings %>% 
  ggplot() +
  geom_boxplot(aes(armed,comp_income)) +
  labs(x="Weapon", y="Houshold Income As Multiple of County Income for Shot Persons", title="The Weapon Used and the Relative Finanical Situation of Those Shot by Police (2015)") +
  theme(plot.title=element_text(hjust=.5, face="bold", size=10),
        axis.title.y = element_text(size=10))
```
  
This graph above shows the weapon of preference for those shot and their financial status relative to the county. All median county income multiples by weapon, with the exception of "unknown", were poorer than their county income. This suggests that police shootings disproportionately occur in economically depressed neighborhoods. There were several outcasts where a firearm user had an income substantially higher than their county. The reason for this is unclear, although it is interesting that firearms has the most outliars. It is possible that, because firearms are legal to carry in most jurisdictions, the demographic of those involved in these shootings are different, in this case the demographic being wealthier individuals. It makes sense that if someone has a firearm on them, they are more likely to get violent, hence more shootings.  
 
```{r fifth, warning=FALSE, message=FALSE, echo = FALSE}
crimeAndPoliceInvolvement %>%
  select(raceethnicity, gender, armed, cause) %>%
  group_by(raceethnicity, gender) %>%
  summarize(total=n(), armed, gender, raceethnicity) %>%
  arrange(gender) %>%
  ggplot(aes(raceethnicity, fill=armed)) +
  geom_bar() + coord_flip() +
  labs(x="Race/Ethnicity", y="Number of People Shot", title="Number of People Shot by Race and Arms (2015)")
``` 
  
This figure shows that whites are shot the most with blacks the second most. Because the populations differ as well as the violent crime rates between the two races, it is not responsible to say that whites or blacks are over represented in those shot and killed by police. The data also shows that there are not major differences in the weapons being carried by those who were shot and killed. This is interesting since blacks usually live in inner cities where weapon restrictions are greater. However, it is possible that most of these interactions occur in similar types of neighborhoods, resulting in similar shooting victim characteristics, regardless of race. 
  
**Conclusion**  
  
In conclusion, these datasets show that there are correlations between violence in states and police shootings. It also shows that mass incarcertation is not taking place in the way that most believe. In terms of police shootings, police shootings show that most of those shot were armed, and therefore, the police probably had a reasonable fear during their interaction with the suspect. Therefore, even thought police brutality claims must be analyzed on a case by case basis, it appears evident that police are not randomly shooting Americans on the average. Moreover, the last graph shows that Blacks are shot substantially less than Whites, although further analysis must be done on if these shootings are proportional to population or crime committed by both races. It also seems that state population has a lot more correlation to police shootings than violence, as California, Texas, and Florida top the list for police shootings, and for highest populations in the country. Further, roughly half of the victims shot in each state were carrying a firearm. This is interesting as some states are more strict about firearm ownership than others, bringing to question if laws restricting carrying a firearm will make a difference in terms of police shootings. For example, California gun laws are severely stricter than Texas and Florida, however, the percentage of suspects with firearms are roughly the same between Florida and California, and in Texas the percentage is significantly lower. Finally, it is clear that the vast majority of those shot and killed by police came from poorer socioeconomic backgrounds, although there were several outliars. This is potentially because folks in poorer areas have higher crime rates, leading to more policing.
Finally, this dataset shows us that the proportion of criminals incarcerated for violent crime increases as states have more prisoners per capita. This suggests that highly incarcerated states are not rounding up petty crimes and sending those individuals to jail. The reason lots of the analysis here states that more analysis must be done is because stating an opinion with confidence from this dataset may lead to disappointment when further analysis is done. Although this gives data scientists a starting ground for analyzing the criminal justice system and policing, there are many variables that the datasets leave out. For example, how many of those shot and killed has a prior arrest for violent crimes. How likely were the police to shoot someone in this area. Were black police officers just as likely to shoot the black suspects as white police officers? All of these are relevant questions in assessing our countries law enforcement and justice system. 
  
**References**
  
1) https://www.kaggle.com/christophercorrea/prisoners-and-crime-in-united-states
  
2) https://www.kaggle.com/mysarahmadbhat/police-killings
  
**Acknowledgments**

1) Bureau of Justice Statistics

2) FBI Uniform Crime Reporting Statistics







 